Planning with Hierarchical Temporal Memory for Deterministic Markov Decision Problem

### 2020-12-19 reports/v1_0/paper_eng/paper_eng.tex
polishing paper text

Reinforcement Learning (RL) seeks to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks. It requires from such methods to learn a policy for sequential decision-making problems, which are commonly formalized as MDP optimization \cite{Puterman_1994}. Recently, RL has achieved notable landmarks in various task domains such as classical board games chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker} and visually rich computer games like Atari \cite{Mnih_2015_Atari} and Dota 2 \cite{openai_2019_dota}. At the core of some of the methods used, especially for board games, were planning algorithms, which rely on the knowledge of the environment dynamics. However, the model of the environment is unknown in general case, and there are two common opposite approaches in RL to deal with it: a model-free and a model-based.

Model-free approach doesn't explicitly learn and make use of the knowledge of the environment. Instead, it directly learns the global policy from interactions with the environment. While model-free approach has many successful examples among its disadvantages are data inefficiency as it requires high amount of interactions with the environment and struggles for precise lookahead \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}.

Model-based RL can address the issues of both planning and model-free RL by first learning the environment dynamics and then to exploit it for more efficient learning of the agent's policy \cite{moerland_2020_modelbased}. Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model. After the model has been learned, MDP planning method, such as MCTS \cite{Coulom_2007_mcts}, can be applied to infer the optimal policy. One model-based RL approach is to build a model that operates on the raw input data level \cite{Gorodetskiy2020,kaiser_2020_modelbased}. For visually rich environments this may be expensive in terms of computations and also may lead to a model attentive to unimportant details. Another approach is to build a more compact latent-space model instead, for example, to represent an abstract MDP equivalent to the real one in terms of state values \cite{silver_2020_muzero}.

To model the environment it's common to use Artificial Neural Networks (ANNs) \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}. But it's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}. Hierarchical Temporal Memory (HTM) framework \cite{George_Hawkins_2009} is an example of such human-like memory model. This framework utilizes discrete binary sparse distributed representations (SDR) \cite{Cui_Ahmad_Hawkins_2017_sdr} that enables more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings. HTM also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM}. Learning in TM isn't based on the gradient descend, and typically it learns with a faster rate compared to common ANNs.

In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics, supplied with the planning algorithm to infer the optimal global policy. Some attempts to extend the HTM model have been made before \cite{Skrynnik2016,Daylidyonok2018,Nugamanov2020}, but the full implementation of the planning subsystem with consideration for reinforcement signal was not implemented in other works. We study applicability of the presented memory model and viability of the planning method in different maze environments.

@@

Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of the sequential memory. It works with sparse distributed representations (SDR), i.e. sparse binary vectors, which are typically high-dimensional. Therefore, the data encoding scheme should be chosen too. In \ref{sec_MemoryModule} we describe how agent's memory is organized on top of TM, so this section mostly covers the TM's input.

In our work both actions and states are represented as integer numbers from a fixed range (although, different for states and actions). Thus, we use a simple encoding scheme, where numbers from some range $[0, N)$ are mapped to non-overlapping equally-sized sets of active bits of the output SDR vector. For example for $N = 3$ numbers 0, 1 and 2 are encoded correspondingly as following:

@@

As you see, resulting vector is split into buckets of bits, and the size of buckets is a hyperparameter (in the example above it's 4).

For SDR vectors a union operation is defined as a bitwise OR applied to the corresponding vectors. Also we define a similarity metric, which is the number of the same one-bits of the vectors in consideration, which is, again, the same as the dot product of such vectors.

@@

We present results for two mazes: \texttt{multi\_way\_v0} and \texttt{multi\_way\_v2} (see Fig.~\ref{fig_comparison_mw} and Fig.~\ref{fig_detailed_mw0}). Each experiment in \texttt{multi\_way\_v0} had 4 different rewarding goal positions being sequentially changed every 100 episodes. Each experiment in \texttt{multi\_way\_v2} had 6 rewarding goal positions being changed every 200 episodes. Schematic view of these mazes is provided in Appendix ~\ref{apx_handcrafted_mazes}.

@@

\texttt{multi\_way\_v2}. Each experiment had 6 rewarding places being changed every 200 episodes:

>>>>

Reinforcement Learning (RL) seeks to find methods that enable robots or virtual agents to act autonomously in order to solve human-defined tasks. It requires from such methods to learn a policy for sequential decision-making problems  which are commonly formalized as MDP optimization \cite{Puterman_1994}. Recently, RL has achieved notable landmarks in various task domains such as classical board games chess \cite{Campbell_Hoane_Hsu_2002_DeepBlue}, Go \cite{Silver_Go}, poker \cite{Brown_Sandholm_2017_Poker} and visually rich computer games like Atari \cite{Mnih_2015_Atari} and Dota 2 \cite{openai_2019_dota}. At the core of some of the methods used, especially for board games, were planning algorithms, which rely on the knowledge of the environment dynamics. However, the model of the environment is unknown in general case, and there are two common opposite approaches in RL to deal with it: a model-free and a model-based.

Model-free approach doesn't explicitly learn and make use of the knowledge of the environment. Instead, it directly learns the global policy from interactions with the environment. While model-free approach has many successful examples, among its disadvantages are data inefficiency as it requires high amount of interactions with the environment and struggles for precise lookahead \cite{Mnih_2015_Atari,haarnoja_2018_sac,schulman_2017_ppo}.

Model-based RL can address the issues of both planning and model-free RL by first learning the environment dynamics and then to exploit it for more efficient learning of the agent's policy \cite{moerland_2020_modelbased}. Typically, the model is represented by MDP and consists of two components: a state transition model and a reward model. After the model has been learned, MDP planning method, such as MCTS \cite{Coulom_2007_mcts}, can be applied to infer the optimal policy. One model-based RL approach is to build model that operates on the raw input data level \cite{Gorodetskiy2020,kaiser_2020_modelbased}. For visually rich environments this may be expensive in terms of computations and also may lead to a model attentive to unimportant details. Another approach is to build a more compact latent-space model instead, for example, to represent an abstract MDP equivalent to the real one in terms of state values \cite{silver_2020_muzero}. While it can help dealing with drawbacks of the former approach, it also requires additional efforts to ensure that sufficiently good latent-space mapping is learned. To model the environment it's common to use Artificial Neural Networks (ANNs) \cite{silver_2020_muzero,Ha_Schmidhuber_2018_worldmodels}. It's also very tempting to take more inspiration from human brains, especially for how our memory functions \cite{Hassabis_2017_neuro}.

Hierarchical Temporal Memory (HTM) framework \cite{George_Hawkins_2009} is an example of such human-like memory model. This framework utilizes discrete binary sparse distributed representations (SDR) that enable more efficient data processing, high noise tolerance and a natural way to set similarity metric between data embeddings \cite{Cui_Ahmad_Hawkins_2017_sdr}. HTM also provides a model of the sequential temporal memory (TM) \cite{hawkins_TM,Ahmad_Lavin_2017_data_streams}. Learning in TM is not based on gradient descend, and typically it learns with a faster rate compared to common ANNs \cite{Cui_Ahmad_Hawkins_2016_seq_learn}.

In our work we present a simplified memory-based framework based on the ideas of HTM that can be used to model the environment dynamics. Great part of our contribution is that we also supply it with the planning algorithm that enables an agent to infer the optimal global policy. Some attempts to extend the HTM model have been made before \cite{Skrynnik2016,Daylidyonok2018,Nugamanov2020}, but the full implementation of the planning subsystem with consideration for reinforcement signal was not implemented in other works. We study applicability of the presented memory model and viability of the planning method in different maze environments.

@@

Temporal Memory (TM) from Hierarchical Temporal Memory (HTM) framework is a model of sequential memory. It works with sparse distributed representations (SDR), i.e. sparse binary vectors, which are typically high-dimensional. In general, an input data is not represented as SDR vectors. Therefore, the data encoding scheme should be chosen too, and this section mostly covers questions related TM's input. Further in \ref{sec_MemoryModule} we describe how agent's memory is organized on top of TM.

In our work both actions and states are represented as integer numbers from a fixed range (although, different for states and for actions). Thus, we use simple encoding scheme, where numbers from a range $[0, N)$ are mapped to non-overlapping equally-sized sets of active bits of the output SDR vector. For example for $N = 3$ numbers from the range $[0, 3)$ are encoded correspondingly as following:

@@

As you see, resulting vector is split into buckets of 4 bits. In general the size of buckets is a hyperparameter (we mostly used 8 or 10).

A union operation is defined for SDR vectors - a bitwise OR applied to the corresponding vectors. We also define similarity metric, which is the number of the same one-bits of the vectors in consideration. Or it can be seen as the dot product of such vectors, which among other things draws parallels between semantics and geometry, e.g. non-related SDR vectors are orthogonal.

@@

We present results for two mazes: \texttt{multi\_way\_v0} and \texttt{multi\_way\_v2} (see Fig.~\ref{fig_comparison_mw} and Fig.~\ref{fig_detailed_mw0}). Each experiment in \texttt{multi\_way\_v0} had 4 different rewarding goal positions being sequentially changed every 100 episodes. Each experiment in \texttt{multi\_way\_v2} had 6 rewarding goal positions being changed every 100 episodes. Schematic view of these mazes is provided in Appendix ~\ref{apx_handcrafted_mazes}.

@@

\texttt{multi\_way\_v2}. Each experiment had 6 rewarding places being changed every 100 episodes:
